---
title: "Diagnosing Pipeline Runs"
author: "Michael Hallquist & Dan Shallal"
date: "25 Feb 2026"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Diagnosing Pipeline Runs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

Running an fMRI pipeline with BrainGnomes typically means launching **many**
dependent jobs: BIDS conversion, MRIQC, fMRIPrep, ICA‑AROMA, one or more
postprocessing streams, and optional ROI extraction. Each job writes its own
logs and completion markers, and jobs depend on one another, often across
multiple runs (sequences) of the same project. Without help, it is hard to
answer even basic questions:

- What already succeeded?
- What is still running or queued?
- What failed, and was it the *root cause* or just downstream of another
  failure?
- Where should I look next in the logs?

BrainGnomes provides three complementary tools for this:

- `get_project_status()` gives a project‑wide table of step‑level completion
  flags and times for all subjects.
- `get_subject_status()` focuses that same information on a single subject
  (and session), making it easy to see where one subject is stuck.
- `diagnose_pipeline()` ties together the job‑tracking database and log paths
  into an interactive tree view for deep, log‑driven debugging.

This vignette walks through a practical workflow that starts with the
high‑level `*_status()` summaries and then uses `diagnose_pipeline()` when you
need to understand *why* something failed and which specific jobs are involved.

The examples below were run on this project:

`/proj/mnhallqlab/projects/bg08test_14feb2026`

## Load the project

```{r, eval=FALSE}
library(BrainGnomes)

project_dir <- "/proj/mnhallqlab/projects/bg08test_14feb2026"
scfg <- load_project(project_dir, validate = FALSE)
```

`validate = FALSE` is useful when loading from non-interactive contexts, because
validation can otherwise trigger interactive correction prompts.

# Project-level triage with `get_project_status()`

Start with a table of completion flags and times for all subjects:

```{r, eval=FALSE}
status_df <- get_project_status(scfg)
status_df
summary(status_df)
```

Captured output (16 Feb 2026):

```
rows: 10 cols: 12

sub_id  bids_conversion_complete  mriqc_complete  fmriprep_complete  aroma_complete  task_complete
540294  TRUE                     TRUE            FALSE              FALSE           FALSE
540295  TRUE                     TRUE            FALSE              FALSE           FALSE
...     ...                      ...             ...                ...             ...
540311  TRUE                     TRUE            FALSE              FALSE           FALSE

Summary:
step                       n_complete
bids_conversion_complete   10
mriqc_complete             10
fmriprep_complete           0
aroma_complete              0
task_complete               0
```

This gives a quick pass/fail style view. In this run, BIDS conversion and MRIQC
are complete for all subjects, while downstream steps are not yet complete.

Under the hood, `get_project_status()` and `get_subject_status()`:

- look in the project’s BIDS directory to discover which subjects/sessions
  exist, and
- scan each subject’s log directory for BrainGnomes `.complete` (and related)
  marker files.

From these filesystem markers, they derive strict `*_complete` flags and
completion times for each enabled step.

You can also scan log markers for explicit failures:

```{r, eval=FALSE}
list.files(
  scfg$metadata$log_directory,
  pattern = "_fail$",
  recursive = TRUE,
  full.names = FALSE
)
```

Captured output:

```
[1] "sub-540295/.fmriprep_sub-540295_fail"
[2] "sub-540300/.fmriprep_sub-540300_fail"
[3] "sub-540301/.fmriprep_sub-540301_fail"
[4] "sub-540302/.fmriprep_sub-540302_fail"
[5] "sub-540303/.aroma_sub-540303_fail"
```

# Subject-level check with `get_subject_status()`

To check one subject:

```{r, eval=FALSE}
get_subject_status(scfg, "540296")
```

Captured output:

```
  sub_id ses_id bids_conversion_complete bids_conversion_time mriqc_complete
1 540296   <NA>                     TRUE  2026-02-14 18:27:01           TRUE
           mriqc_time fmriprep_complete fmriprep_time aroma_complete aroma_time
1 2026-02-15 04:00:02             FALSE          <NA>          FALSE       <NA>
  task_complete task_time
1         FALSE      <NA>
```

This is often the fastest way to answer, "Where is subject X stuck?"

# Interactive diagnosis with `diagnose_pipeline()`

Running a full fMRI preprocessing + postprocessing pipeline generates many jobs,
log files, and potential failure points. Manually chasing SLURM job IDs and log
files quickly becomes unmanageable. The goal of `diagnose_pipeline()` is to
wrap all of this job-tracking information into a single, interactive view.

Behind the scenes, every job submitted by `run_project()` is recorded in an
SQLite database associated with the project. Each record stores:

- the **sequence ID** (a particular end‑to‑end run of the pipeline),
- the **subject**, job name, scheduler ID, and parent/child relationships,
- timestamps (submitted, started, finished),
- scheduler options and exit codes,
- and a derived **status** (`COMPLETED`, `STARTED`, `QUEUED`, `FAILED`,
  `FAILED_BY_EXT`).

`diagnose_pipeline()` reads this SQLite database, reconstructs the job
dependency tree, and then guides you through it interactively. The function
returns that raw tree structure invisibly so you can also inspect it yourself:

```{r, eval=FALSE}
tree <- diagnose_pipeline(scfg)
print(tree)
```

Each node in `tree` corresponds to a single tracked job and includes its
subject, status, scheduler metadata, and parent/child links. This is the same
structure BrainGnomes uses internally to understand job dependencies.

## Sequence‑level diagnosis

The first prompt asks whether to diagnose by **subject** or by **sequence ID**:

```
Enter 1 for subject summary or 2 for sequence ID
> 2

Enter which pipeline run to diagnose. The default is the most recent.
> [Enter]
```

Choosing the sequence‑first path lets you pick a specific pipeline run (for
example, if you re‑ran the project with a different configuration). Once you
select a sequence, `diagnose_pipeline()` shows all **top‑level jobs** for that
run, grouped by subject, with their current status:

```
Postprocess stream1
  postprocess_stream1_sub-540294                  [STARTED]
  ---postprocess_sub-540294_task-ridl_run-01_...   [FAILED]
```

For postprocessing, this view expands nested structure: the subject‑level
postprocess job has **child jobs** for each postprocess stream, and those in
turn have children for each image‑level job (e.g., individual BOLD runs).
At a glance you can see which specific image(s) caused a stream or subject to
fail.

From the sequence‑level view you can choose a job and drill down:

- show a detailed summary (times, parent/child jobs, exit codes),
- open the associated stdout or stderr log in the console,
- or return the log contents as a character vector for further inspection.

This is usually the fastest way to answer, *“Why did this particular run of the
pipeline fail?”* and *“Which exact job needs debugging?”*

## Subject‑level diagnosis across sequences

If you already know which subject you care about, you can start with the
subject‑summary branch:

```
Enter 1 for subject summary or 2 for sequence ID
> 1

Found 10 subject(s) across all runs.
Enter the number corresponding to the subject you want to view
> 3
```

Here, `diagnose_pipeline()` looks across **all sequences** in the SQLite
database and shows, for the chosen subject, which combination of runs produced
the best overall completion pattern. This helps with questions like:

- Did subject `sub-540303` ever complete fMRIPrep + AROMA + postprocessing?
- If so, which sequence (which configuration / run) produced that success?

The summary groups jobs by step (BIDS conversion, MRIQC, fMRIPrep, AROMA,
postprocessing streams, etc.) and highlights, for each step, which sequence
had the most successful outcome. You can then jump from this high‑level
summary back into sequence‑ or job‑level inspection for detailed debugging.

Example subject‑level summary for `sub-540303`:

```
BIDS Conversion
  ✓ bids_conversion_sub-540303 (job 15135998, sequence 18303423-c037-477a-91af-b6289525dad2) [COMPLETED]

fMRIPrep
  • fmriprep_sub-540303 (job 15136000, sequence 18303423-c037-477a-91af-b6289525dad2) [STARTED]
```

## Interpreting job statuses

`diagnose_pipeline()` uses the following job‑level statuses:

- `COMPLETED`: the job finished successfully (based on scheduler status and
  BrainGnomes tracking).
- `STARTED`: the job has begun running or previously ran without a terminal
  status yet recorded.
- `QUEUED`: the scheduler knows about the job, but it is waiting on
  dependencies or resources.
- `FAILED`: the job itself failed (non‑zero exit code or explicit failure
  marker).
- `FAILED_BY_EXT`: the job was never allowed to run successfully because an
  upstream dependency failed. For example, if one image‑level postprocess job
  in a stream fails, its parent stream‑level job and other dependent jobs are
  marked `FAILED_BY_EXT`.

This distinction is important: `FAILED_BY_EXT` tells you which jobs are
*victims* of an upstream problem versus the **true root cause** that needs
debugging.

In contrast, `get_project_status()` and `get_subject_status()` report simple
`*_complete` flags per step; these are stricter than raw scheduler status and
rely on BrainGnomes completion markers and expected outputs. A common workflow
is:

1. Use `get_project_status()` or `get_subject_status()` for a quick overview of
   which steps are done. These functions do **not** consult the SQLite database;
   instead they scan the project’s BIDS and log directories for `.complete`
   marker files and expected outputs, then derive strict `*_complete` flags and
   completion times per step.
2. Use `diagnose_pipeline()` when you need to chase down *why* a particular
   job or subject failed and which logs to read next. This function *does* read
   from the SQLite job‑tracking database in order to reconstruct the full job
   dependency tree and status history across all runs.
